---
title: "Assignment 2 Part 3 - Language Development in ASD - Power and simulations"
author: "[YOUR NAME]"
date: "[DATE]"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this part of the assignment, we try to figure out how a new study should be planned (i.e. how many participants?) in order to have enough power to replicate the findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8):
1- if we trust the estimates of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
2- if we are skeptical of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
3- if we only have access to 30 participants. Identify the power for each relevant effect and discuss whether it's worth to run the study and why
The list above is also what you should discuss in your code-less report.


## Learning objectives

- Learn how to calculate statistical power
- Critically appraise how to apply frequentist statistical power

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- Load your dataset (both training and testing), fit your favorite model, assess power for your effects of interest (probably your interactions).
- Report the power analysis and comment on what you can (or cannot) use its estimates for.
- Test how many participants you would have to have to replicate the findings (assuming the findings are correct)

N.B. Remember that main effects are tricky once you have interactions in the model (same for 2-way interactions w 3-way interactions in the model). If you want to test the power of main effects, run a model excluding the interactions.
N.B. Check this paper: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12504
You will be using:
- powerSim() to calculate power
- powerCurve() to estimate the needed number of participants
- extend() to simulate more participants

```{r}
#Loading packages

#library(githubinstall)
#githubinstall("simr", lib = .libPaths())
#print(pacman::p_path() == .libPaths())## [1] TRUE TRUE

pacman::p_load(pacman, tidyverse, lmerTest, readr,dplyr,stringr,lmerTest,Metrics,caret, merTools, groupdata2, simr)


#Data loading, Kenneth Style


CleanUpData <- function(Demo,LU,Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      Child.ID = SUBJ, 
      Visit = VISIT) %>%
    mutate(
      Visit = as.numeric(str_extract(Visit, "\\d")),
      Child.ID = gsub("\\.","", Child.ID)
      ) %>%
    dplyr::select(
      Child.ID, Visit, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    dplyr::select(
      Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      Child.ID = gsub("\\.","", Child.ID)
    )
    
  Data = merge(Demo, Speech, all = T)
  
  Data1 = Data %>% 
    subset(Visit == "1") %>% 
     dplyr::select(Child.ID, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(Ados1 = ADOS, 
            verbalIQ1 = ExpressiveLangRaw, 
            nonVerbalIQ1 = MullenRaw,
            Socialization1 = Socialization) 
  
  Data = merge(Data, Data1, all = T) %>%
    mutate(
      Child.ID = as.numeric(as.factor(as.character(Child.ID))),
      Visit = as.numeric(as.character(Visit)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "ASD", # Note that this function is fixed here
         "B"  = "TD")
    )

  return(Data)
}

# Training Data
#setwd('~/Dropbox/2019 - methods 3/Assignments19/Assignment2/solutions/')
Demo <- read_csv('demo_train.csv')
LU <- read_csv('LU_train.csv')
Word <- read_csv('token_train.csv')

TrainData <- CleanUpData(Demo,LU,Word)

Demo <- read_csv('demo_test.csv')
LU <- read_csv('LU_test.csv')
Word <- read_csv('token_test.csv')

TestData <- CleanUpData(Demo,LU,Word)

# merge training and testing
Data <- merge(TrainData, TestData, all = T)

Data <- Data[complete.cases(Data[,c("CHI_MLU","Visit","Diagnosis","verbalIQ1","Child.ID")]),]
Data$Child.ID <- as.factor(Data$Child.ID)

#The way Kenneth does it
model_K <- lmer(CHI_MLU ~ Visit * Diagnosis + (1+Visit|Child.ID), Data, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE))

sim = powerSim(model_K , fixed("Diagnosis"), nsim = 50, seed = 1, progress = F) 
sim


fixef(model_K)["DiagnosisTD"] <- 0.4

sim2 = powerSim(model_K , fixed("Diagnosis"), nsim = 50, seed = 1, progress = F) 
sim2

#Now for visit:

sim3 = powerSim(model_K, fixed("Visit"), nsim = 50, seed = 1, progress = F)
sim3


    ## Power for predictor 'Visit', (95% confidence interval):
    ##        0.00% ( 0.00,  7.11)
    ## 
    ## Test: unknown test
    ##       Effect size for Visit is 0.13
    ## 
    ## Based on 50 simulations, (50 warnings, 50 errors)
    ## alpha = 0.05, nrow = 387
    ## 
    ## Time elapsed: 0 h 0 m 4 s
    ## 
    ## nb: result might be an observed power calculation

#**50 errors**, that is a problem (if there is any errors don’t worry about the power). Lets examine:


print(sim3$errors$message[1])


    ## [1] "Models have either equal fixed mean stucture or are not nested"
print(sim3$warnings$message[1])

    ## [1] "Main effect (Visit) was tested but there were interactions."

##Holy crap, all it is complaining about is the presence of an interaction, and here I spent hours trying to figure out why it didn't work. Bad function structure indeed.


#And now for the interaction
fixef(model_K)["Visit:DiagnosisTD"] <- 0.1 # let's try setting a fixed ef
powerSim(model_K, fixed("Visit:Diagnosis"), nsim = 50, seed = 1, progress = F)

#Hmm, power of 87%. Not bad. But good enough?

model_K <- extend(model_K, along = "Child.ID", n = 120) #extend data along child ID

# plot the powercurve
power_curveV1 = powerCurve(model_K, fixed("Visit:Diagnosis"), along = "Child.ID", 
                  nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F) # waaay to few sim
# break is a which interval is should do a power calculations (this simply says every 5th child)
plot(power_curveV1)

#Contrary to Kenneth again, I find that our participant number is fine and dandy. I think we need to run more sims.



#Loading the csv's of the data cleaned last time

power_train <- read.csv("Cleaned Training Data.csv")
power_test <- read.csv("Cleaned Test Data.csv")

all_power <- rbind(power_train, power_test)

all_power$Child.ID <- as.factor(all_power$Child.ID)

#The way Kenneth does it (with my own dataframe added)
interact_mod <- lmer(CHI_MLU ~ Visit * Diagnosis + (1+Visit|Child.ID), all_power, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE))








sim = powerSim(interact_mod, fixed("Diagnosis"), nsim = 50, progress = F)
sim
#I get widely different power results, but at least I get results. Alright, let's try with Visit instead.

sim2 = powerSim(interact_mod, fixed("Visit"), nsim = 50, progress = F)
sim2



#Ensuring Child.ID is the right type of variable


#Removing incomplete cases:
unlimited_power <- unlimited_power[complete.cases(unlimited_power[,c("CHI_MLU", "Visit", "Diagnosis", "verbalIQ1", "Child.ID")]),]

#- Fitting our favourite mode
power_model <- lmer(CHI_MLU ~ Visit + (1|Child.ID) + (0 + Visit|Child.ID), unlimited_power, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE))
summary(power_model)

fixef(power_model)["Visit"] <- 0.3

powerSim(power_model, fixed("Visit"), nsim = 50, progress = F)


```


### Exercise 2

How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
- assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
- if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.

```{r}

```


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}

```
