---
title: "Assignment 2 Part 3 - Language Development in ASD - Power and simulations"
author: "Asger, Rebecca, Sophia, Jana, Dora"
date: "04-10-2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this part of the assignment, we try to figure out how a new study should be planned (i.e. how many participants?) in order to have enough power to replicate the findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8):
1- if we trust the estimates of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
2- if we are skeptical of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
3- if we only have access to 30 participants. Identify the power for each relevant effect and discuss whether it's worth to run the study and why
The list above is also what you should discuss in your code-less report.


## Learning objectives

- Learn how to calculate statistical power
- Critically appraise how to apply frequentist statistical power

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- Load your dataset (both training and testing), fit your favorite model, assess power for your effects of interest (probably your interactions).
- Report the power analysis and comment on what you can (or cannot) use its estimates for.
- Test how many participants you would have to have to replicate the findings (assuming the findings are correct)

N.B. Remember that main effects are tricky once you have interactions in the model (same for 2-way interactions w 3-way interactions in the model). If you want to test the power of main effects, run a model excluding the interactions.
N.B. Check this paper: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12504
You will be using:
- powerSim() to calculate power
- powerCurve() to estimate the needed number of participants
- extend() to simulate more participants

```{r}
#Loading packages

#library(githubinstall)
#githubinstall("simr", lib = .libPaths())
#print(pacman::p_path() == .libPaths())## [1] TRUE TRUE

pacman::p_load(pacman, tidyverse, lmerTest, readr,dplyr,stringr,lmerTest,Metrics,caret, merTools, groupdata2, simr)


# Data loading, Riccardo Style!


CleanUpData <- function(Demo,LU,Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      Child.ID = SUBJ, 
      Visit = VISIT) %>%
    mutate(
      Visit = as.numeric(str_extract(Visit, "\\d")),
      Child.ID = gsub("\\.","", Child.ID)
      ) %>%
    dplyr::select(
      Child.ID, Visit, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    dplyr::select(
      Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      Child.ID = gsub("\\.","", Child.ID)
    )
    
  Data = merge(Demo, Speech, all = T)
  
  Data1 = Data %>% 
    subset(Visit == "1") %>% 
     dplyr::select(Child.ID, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(Ados1 = ADOS, 
            verbalIQ1 = ExpressiveLangRaw, 
            nonVerbalIQ1 = MullenRaw,
            Socialization1 = Socialization) 
  
  Data = merge(Data, Data1, all = T) %>%
    mutate(
      Child.ID = as.numeric(as.factor(as.character(Child.ID))),
      Visit = as.numeric(as.character(Visit)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "ASD", # Note that this function is fixed here
         "B"  = "TD")
    )

  return(Data)
}

# Loading the Training Data
Demo <- read_csv('demo_train.csv')
LU <- read_csv('LU_train.csv')
Word <- read_csv('token_train.csv')

train_data <- CleanUpData(Demo,LU,Word)

# Loading the Test Data
Demo <- read_csv('demo_test.csv')
LU <- read_csv('LU_test.csv')
Word <- read_csv('token_test.csv')

test_data <- CleanUpData(Demo,LU,Word)

# Merging training and testing data
data <- merge(train_data, test_data, all = T)

# Removing the NA's/incomplete cases
data <- data[complete.cases(data[,c("CHI_MLU","Visit","Diagnosis","verbalIQ1","Child.ID")]),]

# Turning Child.ID into a factor
data$Child.ID <- as.factor(data$Child.ID)

# Defining our favourite model.
fav <- lmer(CHI_MLU ~ Visit*Diagnosis + verbalIQ1 + MOT_MLU + (1|Child.ID) + (0 + Visit|Child.ID), data = data, REML = F, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE, optCtrl = list(ftol_abs = 1e-10, xtol_abs = 1e-10, maxeval=10000)))
summary(fav) # Everything is still significant, wonderful and bliss.

# Simulating the power of our interaction effect

# First we evaluate power for our interaction effect.
favsim = powerSim(fav, fixed("Visit:Diagnosis"), nsim = 50, seed = 1, progress = F) 
favsim # Power of 100%! Of course, this is an observed power calculations.

# And now for verbal IQ at visit 1
vIQsim = powerSim(fav, fixed("verbalIQ1"), nsim = 50, seed = 1, progress = F)
vIQsim #100% power!

# Mother's MLU
momsim = powerSim(fav, fixed("verbalIQ1"), nsim = 50, seed = 1, progress = F)
momsim #100% Power!


fav <- extend(fav, along = "Child.ID", n = 120) #extend data along child ID to simulate 120 participants.

# plot the powercurve
power_curve = powerCurve(fav, fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)

plot(power_curve)  # Interestingly, we get power of 100% even for 10 participants. Again, observed power calculations.

```


### Exercise 2

How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
- assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
- if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.

```{r}

# First we specify the smallest effect of interest of the interaction between visit and diagnosis.

fixef(fav)["Visit:DiagnosisTD"] <- 0.5 # we set the fixed effect to 0.5; a difference of half a morpheme across time and diagnosis seems to be worth paying attention to.
favsim = powerSim(fav, fixed("Visit:Diagnosis"), nsim = 50, seed = 1, progress = F) 
favsim # Again, we get power of 100%. Which might not be that informative since, technically, the effect size we got for the interaction effect was exactly 0.5. (We agreed on a 0.5 effect size before actually looking at what the effect size was).


fixef(fav)["Visit:DiagnosisTD"] <- 0.05 # Just for fun (since an effect of 0.5 gives us very high power), let's try setting the fixed effect to 0.05.
notfavsim = powerSim(fav, fixed("Visit:Diagnosis"), nsim = 50, seed = 1, progress = F) 
notfavsim # Power of 41.10%. In other words, very low power indeed.


fav <- extend(fav, along = "Child.ID", n = 120) #extend data along child ID to simulate 120 participants.

# plot the powercurve of the interaction, for an effect size of 0.05
power_curve = powerCurve(fav, fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)

plot(power_curve)  
power_curve # Roundabouts 85 participants would be necessary to detect an effect size of such small magnitude 80% of the time.



#Now for verbal IQ at visit 1:
fixef(fav)["verbalIQ1"] <- 0.1 # We expect verbal IQ at visit 1 to account for a less dramatic difference in the kids MLU, than Visit and Diagnosis.
vIQsim = powerSim(fav, fixed("verbalIQ1"), nsim = 50, seed = 1, progress = F)
vIQsim #100% powaaaa!


# We make the power curve and plot it
IQ_curve = powerCurve(fav, fixed("verbalIQ1"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)
plot(IQ_curve)
IQ_curve
# 100% power with 10 participants? Either we are doing something wrong or the effect is very robust (which makes sense; verbal IQ and Child MLU seem to measure the same).


# And now for the mother's MLU
fixef(fav)["MOT_MLU"] <- 0.1 # We likewise expect mother's MLU at visit 1 to account for a less dramatic difference in the kids MLU, than Visit and Diagnosis.
momsim = powerSim(fav, fixed("MOT_MLU"), nsim = 50, seed = 1, progress = F)
momsim #80% Power! [Probably needs more sims]

# We make the power curve and plot it
mom_curve = powerCurve(fav, fixed("MOT_MLU"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)
plot(mom_curve)
mom_curve
# We need more participants than 120 to get a proper estimate power.


fav <- extend(fav, along = "Child.ID", n = 150) #extend data along child ID to simulate 300 participants.

# plot the powercurve
mom_curve2 = powerCurve(fav, fixed("MOT_MLU"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 150, by = 5), seed = 1, progress = F)
mom_curve2
plot(mom_curve2)  # Interestingly, we get power of 100% even for 10 participants. That is quite probably a little extreme, but then again, 0.5 is a big effect.

```


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}
# Redoing all the power analyses with 30 participants.

fixef(fav)["Visit:DiagnosisTD"] <- 0.05
int_30_curve = powerCurve(fav, fixed("Visit:Diagnosis"), along = "Child.ID", nsim = 10, breaks = seq(from = 3, to = 30, by = 3), seed = 1, progress = F)

int_30_curve
plot(int_30_curve)

#Now for verbal IQ at visit 1:
fixef(fav)["verbalIQ1"] <- 0.1
vIQ_30_curve = powerCurve(fav, fixed("verbalIQ1"), along = "Child.ID", nsim = 10, breaks = seq(from = 3, to = 30, by = 3), seed = 1, progress = F)
vIQ_30_curve
plot(vIQ_30_curve)

vIQsim #100% powaaaa!


# We make the power curve and plot it
IQ_curve = powerCurve(fav, fixed("verbalIQ1"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)
plot(IQ_curve)
IQ_curve
# 100% power with 10 participants? Either we are doing something wrong or the effect is very robust (which makes sense; verbal IQ and Child MLU seem to measure the same).


# And now for the mother's MLU
fixef(fav)["MOT_MLU"] <- 0.1 # We likewise expect mother's MLU at visit 1 to account for a less dramatic difference in the kids MLU, than Visit and Diagnosis.
momsim = powerSim(fav, fixed("MOT_MLU"), nsim = 50, seed = 1, progress = F)
momsim #80% Power! [Probably needs more sims]

# We make the power curve and plot it
mom_curve = powerCurve(fav, fixed("MOT_MLU"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 120, by = 5), seed = 1, progress = F)
plot(mom_curve)
mom_curve
# We need more participants than 120 to get a proper estimate power.


fav <- extend(fav, along = "Child.ID", n = 150) #extend data along child ID to simulate 300 participants.

# plot the powercurve
mom_curve2 = powerCurve(fav, fixed("MOT_MLU"), along = "Child.ID", nsim = 10, breaks = seq(from = 10, to = 150, by = 5), seed = 1, progress = F)
mom_curve2
plot(mom_curve2)  # Interestingly, we get power of 100% even for 10 participants. That is quite probably a little extreme, but then again, 0.5 is a big effect.
```
