---
title: "Assignment 2 - Language Development in ASD - Making predictions"
author: "Asger, Sophia, Rebecca, Jana, Dora"
date: "2 okt 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Welcome to the second exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time against some new data. Does the model generalize well?
Then we will learn to do better by cross-validating models and systematically compare them.

The questions to be answered (in a separate document) are:
1- Discuss the differences in performance of your model in training and testing data
2- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
3- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

## Learning objectives

- Critically appraise the predictive framework (contrasted to the explanatory framework)
- Learn the basics of machine learning workflows: training/testing, cross-validation, feature selections

## Let's go

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=1
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=1
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=1

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())


```{r, include = FALSE}
# Loading packages in a most peculiar way
pacman::p_load(pacman, readr,dplyr,stringr,lmerTest,Metrics,caret, merTools, groupdata2)


## Clean up function
CleanUpData <- function(Demo,LU,Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      Child.ID = SUBJ, 
      Visit=VISIT) %>%
    mutate(
      Visit = as.numeric(str_extract(Visit, "\\d")),
      Child.ID = gsub("\\.","", Child.ID) #The counterslash-counterslash is because dots mean "everything" in regular expressions.
      ) %>%
    dplyr::select(
      Child.ID, Visit, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    dplyr::select(
      Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      Child.ID = gsub("\\.","", Child.ID)
    )
    
  Data=merge(Demo,Speech,all=T)
  
  Data1= Data %>% 
     subset(Visit=="1") %>% 
     dplyr::select(Child.ID, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(Ados1 = ADOS, 
            verbalIQ1 = ExpressiveLangRaw, 
            nonVerbalIQ1 = MullenRaw,
            Socialization1 = Socialization) 
  
  Data=merge(Data, Data1, all=T) %>%
    mutate(
      Child.ID = as.numeric(as.factor(as.character(Child.ID))),
      Visit = as.numeric(as.character(Visit)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "ASD",
         "B"  = "TD")
    )
  Data = subset(Data, !is.na(CHI_MLU))
  return(Data)
}

##- Loading and cleaning up training data:

Demo_train <- read.csv("demo_train.csv")
LU_train <- read.csv("LU_train.csv")
token_train <- read.csv("token_train.csv")

data_train <- CleanUpData(Demo = Demo_train, LU = LU_train, Word = token_train)

#Writing a csv file with the cleaned data
write.csv(data_train, "Cleaned Training Data.csv")

##- Loading and cleaning up test data:

Demo_test <- read.csv("demo_test.csv")
LU_test <- read.csv("LU_test.csv")
tokens_test <- read.csv("token_test.csv")

data_test <- CleanUpData(Demo = Demo_test, LU = LU_test, Word = tokens_test)

write.csv(data_test, "Cleaned Test Data.csv")


##- For future reference, the above can be skipped if necessary, and the following code used instead (now that we have csv's of the clean data)
#data_train <- read.csv("Cleaned Training Data.csv")
#data_test <- read.csv("Cleaned Test Data.csv")

#- recreate the models you chose last time (just write the code again and apply it to Train Data)

trained_mod <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + verbalIQ1 + MOT_MLU + (1|Child.ID) + (0 + Visit|Child.ID), data = data_train, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE,  optCtrl = list(ftol_abs = 1e-10, xtol_abs = 1e-10, maxeval=10000)))

# Summarizing the model
summary(trained_mod) # Everything is still significant.


#- calculate performance of the model on the training data: root mean square error is a good measure.

train_predictions <- predict(trained_mod)

rmse(data_train$CHI_MLU, train_predictions) # 0.35 root mean square error.


#- test the performance of the models on the test data (Tips: google the functions "predict()")

test_predictions <- predict(trained_mod, data_test)

rmse(data_test$CHI_MLU, test_predictions) # 0.49, i.e. close to 50%. Ouch.
```

The RMSE jumps from 0.351 to 0.493 when the model of children's MLU is applied, respectively, to the train data versus the test data.

### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model (Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)


- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data


```{r}
##- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).

# Creating the basic model:

basic_model <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + (1|Child.ID) + (0 + Visit|Child.ID), data = data_train)

#- Make a cross-validated version of the model;  loop through each fold, train a model on the other folds and test it on the fold.


#Making folds using the fold() function of the groupdata2 package

data_train <- fold(data_train, k = 5, cat_col = "Diagnosis", id_col = "Child.ID") %>% arrange(.folds)

#Since fold() doesn't always fold the same way, we elected to save our first iteration of the folded dataframe as "Folded Data 1", using the code:

#write.csv(data_train, "Folded Data 1.csv")

## That line of code is text now. If we did it again, it would ruin the whole point.

#We now retrieve that original folded dataframe, which we will use as much as possible from now on:
folded_1 <- read.csv("Folded Data 1.csv")

#We can now use the folded dataframe in a for loop to cross validate our basic model:

# First we make a vector to house the RMSE scores the for loop will generate:
RMSE <- numeric(5)

#For loop for the basic model
for(i in 1:5){
  #Make two subsets:
  test_df <- folded_1 %>%  subset(.folds == i)
  train_df <- folded_1 %>%  subset(.folds != i)
  # make a model where you regress the train_df
  mod <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + (1|Child.ID) + (0 + Visit|Child.ID), data = train_df)
  #predict "test_df" based on "mod"
  prediction <- predict(mod, test_df, allow.new.levels = TRUE)
  RMSE[i] <- rmse(test_df$CHI_MLU, prediction)
}


#we got a mean score of
mean(RMSE)
#0.7569 RMSE... Which is not good. But after all, this was the basic model.

#For loop for the hopefully better model (Again, with the original folded dataframe)

RMSE2 <- numeric(5)
for(i in 1:5){
  #Make two subsets:
  test_df <- folded_1 %>%  subset(.folds == i)
  train_df <- folded_1 %>%  subset(.folds != i)
  # make a model where you regress the train_df
  mod <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + verbalIQ1 + MOT_MLU + (1|Child.ID) + (0 + Visit|Child.ID), data = train_df, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE,  optCtrl = list(ftol_abs = 1e-10, xtol_abs = 1e-10, maxeval=10000)))
  #predict "test_df" based on "mod"
  prediction <- predict(mod, test_df, allow.new.levels = TRUE)
  RMSE2[i] <- rmse(test_df$CHI_MLU, prediction)
}

#Convergence issues fixed.
mean(RMSE2)

#A mean RMSE of .541. Still not great, but better at least. By roundabouts 20% or something. Improvement was made!

#- Report the results and comment on them.

#- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

##========
##========< NOT FINISHED YET 

# Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?

## Challenge accepted: More for-loops!
nr_folds <- numeric(9)
mean_rmse <- numeric(9)

#This time, of course, we will have to generate a lot of folded dataframes, so the results will sadly be different each time you run the code. Doing otherwise would require saving 9 dataframes with different foldings, which we didn't do; this is just an exploratory exercise after all.

for(i in 2:10){
  data_train <- read.csv("Cleaned Training Data.csv")
  data_train <- fold(data_train, k = i, cat_col = "Diagnosis", id_col = "Child.ID") %>% arrange(.folds)
  nr_folds[i-1] <- i
  RMSE_plot <- numeric(i)
    for (j in 1:i){
    #Make two subsets:
    test_df <- data_train %>%  subset(.folds == j)
    train_df <- data_train %>%  subset(.folds != j)
    # make a model where you regress the train_df
    mod <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + verbalIQ1 + MOT_MLU + (1|Child.ID) + (0 + Visit|Child.ID), data = train_df,control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE,  optCtrl = list(ftol_abs = 1e-10, xtol_abs = 1e-10, maxeval=10000)))
    #predict "test_df" based on "mod"
    prediction <- predict(mod, test_df, allow.new.levels = TRUE)
    RMSE_plot[j] <- rmse(test_df$CHI_MLU, prediction)
    

    }
  mean_rmse[i-1] <- mean(RMSE_plot)
}

# Saving the number of folds and mean RMSE into a dataframe
plot_df <- data.frame(nr_folds, mean_rmse)
View(plot_df)

# Examining the range of mean RMSE
range(plot_df$mean_rmse)

# Plotting RMSE as a function of number of folds
folds_plot <- ggplot(plot_df, aes(nr_folds, mean_rmse, colour = nr_folds)); folds_plot + geom_point()

# Fitting a linear regression to the data with mean RMSE as a funtion of the number of folds
folds_rmse_lm <- lm(mean_rmse ~ nr_folds)
summary(folds_rmse_lm)
# A cursory linear analysis (Where, again, results will vary every time you run the code) *generally* doesn't show a significant relationship between amount of folds and RMSE.

# Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data

# Well, the original cross-validated error up there using the original folded data was .55, whereas when we tested it earlier it was .49. Similar, not too big of a difference, and by all means too large. Perhaps we should look at better models?

```

Initial 5-fold cross-validation of our model provide a mean RMSE of .541, which obviously is not ideal, but is fairly close to the root mean square error we got when we predicted the test data earlier. In other words, it seems like the RMSE estimate of the cross validation is reliable (and that our model doesn't generalize well).

Also, it seems that adding more folds has a negligible effect on mean RMSE (it still fluctuates around .55)

### Exercise 3) Assessing the single child

Let's get down to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
# So, we will want to extract Bernie specifically. We want a subset of the data with just Bernie.
# First job is to find him though. He shoud be in the object "Demo_test" somewhere:

View(Demo_test)
View(data_test)
# Some eyeballing of these two reveals that Bernie is a TD kid with an ADOS score at visit 1 of 7, which marks him as Child.ID = 2 in the test dataset.

# So we want a subset of the test data that only includes him. We will name this dataframe 'bernie'

bernie <- data_test %>% subset(Child.ID == 2)


#And then we want average MLU scores for the other TD kids in the test set for each visit.
#First we make a dataframe with only TD kids
only_TD <- data_train %>% subset(Diagnosis == "TD")

#Then we group by Visit and summarize mean CHI_MLU
TD_vst_avr <- only_TD %>% group_by(Visit) %>% summarize(mean(CHI_MLU))
#We rename the columns because
colnames(TD_vst_avr) <- c("Visit", "avr_MLU")

#We calculate absolute

absol_dist <- bernie$CHI_MLU - TD_vst_avr$avr_MLU

absol_dist #Bernie is smart.

#Second job is to test our favorite model on Bernie.
trained_model <- lmerTest::lmer(CHI_MLU ~ Visit*Diagnosis + verbalIQ1 + MOT_MLU + (1|Child.ID) + (0 + Visit|Child.ID), data = data_train, control = lmerControl(optimizer = "nloptwrap", calc.derivs = FALSE,  optCtrl = list(ftol_abs = 1e-10, xtol_abs = 1e-10, maxeval=10000)))

#And make some predictions
bernie_predictions <- predict(trained_model, bernie)

error_bernie <- bernie_predictions - bernie$CHI_MLU

error_bernie
# The prediction at the 6th visit seems to be very close to Bernie's actual score (if a little higher). The difference between our predictions and Bernie's scores is smaller than the difference between Bernie and the other TD kids.

#Turn all that into a neat dataframe of predictions and errors.
bernie <- bernie %>% cbind(bernie_predictions, error_bernie)

```

In terms of absolute distance from the mean of his group, Bernie appears to be more talktative than the average TD kid in the train dataset. All of the derived distances were positive (Bernie is above average at each visit).
The predictions made by our model for Bernie's linguistic performance at the 6th visit is very close to Bernie's actual score (only .003 morphemes off prediction, which is essentially nothing).
The difference between our predictions and Bernie's scores is smaller than the difference between Bernie and the other TD kids, indicating that our model (at least in Bernie's case) is not all bad.

### OPTIONAL: Exercise 4) Model Selection via Information Criteria.
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

