---
title: "Assignment 2 - Language Development in ASD - Part 1 - Explaining development"
author: "Asger"
date: "19-09-2019"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
```

# Assignment 2 -  Part 1 - Explaining development

In this assignment you will have to discuss a few important questions (given the data you have). More details below. The assignment submitted to the teachers consists of:
- a report answering and discussing the questions (so we can assess your conceptual understanding and ability to explain and critically reflect)
- a link to a git repository with all the code (so we can assess your code)

Part 1 - Basic description of language development
- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
- Include individual differences in your model of language development (in children). Identify the best model.

Part 2 - Model comparison
- Discuss the differences in performance of your model in training and testing data
- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

Part 3 - Simulations to plan a new study
- Report and discuss a power analyses identifying how many new kids you would need to replicate the results

The following involves only Part 1.

## Learning objectives

- Summarize and report data and models
- Critically apply mixed effects (or multilevel) models
- Explore the issues involved in feature selection


# Quick recap
Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail:
i) relying on actual naturalistic language production,  ii) over extended periods of time.

We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

This RMarkdown file includes 
1) questions (see above). Questions have to be answered/discussed in a separate document that you have to directly send to the teachers.
2) A break down of the questions into a guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results in the doc for the teachers.

REMEMBER that you will have to have a github repository for the code and send the answers to Kenneth and Riccardo without code (but a link to your github/gitlab repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

Before we get going, here is a reminder of the issues you will have to discuss in your report:

1- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
2- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
3- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
4- Include individual differences in your model of language development (in children). Identify the best model.

# Let's go

### Loading the relevant libraries

Load necessary libraries : what will you need?
- e.g. something to deal with the data
- e.g. mixed effects models
- e.g. something to plot with

```{r Load Libraries, include = FALSE}
library(pacman)
p_load(tidyverse)
p_load(lme4, lmerTest, stringi, stringr)

```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data, include = FALSE}
getwd() #The working directory is set when making the project, but just in case I need to change it/use this script in another project:
setwd("C:/Users/Asger/Desktop/Cognitive Science BA/3. Semester/ExpMeth 3/Assignments/Assignment-2")

#Thus, working directory is established. We load the data from last exercise into a dataframe, and call that dataframe "All_data"

all_data <- read.csv("ASD_data.csv")

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Socialization, Visit, Number of words used, Number of unique words used, mean length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}
#We have two tasks:
# 1. Describe your sample (n, age, gender, clinical and cognitive features of the two groups)
# 2. Critically assess whether the groups (ASD and TD) are balanced

#First, we take a look at the data:
View(all_data)
#It is big, confusing and messy. For instance, it has a needless X row. We remove that:
all_data[,1] <- NULL
#Also, child_ID has to be a factor or a character, not numeric:
all_data$child_ID <- as.factor(all_data$child_ID)



#Moreover, the number of participants is inconsistent with the number of observations, as each participant has several observations. To combat this, we can create a new dataframe containing only the observations (using filter() function) from the first visit. We will call this ASD_1st:

first_visit <- all_data %>% filter(visit == 1)


#We can now use the summary() function on ASD_1st to get some of the demographic data:
summary(first_visit)
#This gives us the n of each level of the categorical variable in the dataframe. It also provides us with mean scores of the numerical variables, but since we just removed most of the data, these are uninteresting at the moment. We can infer the following:

#NUMBER AND GENDER: We have 61 participants, 10 Female and 51 Male. This is not surprising; ASD is more common in males.

#ETHNICITY: The majority (53) of participants are white, with African American (2) and White/Latino (2) sharing 2nd place. All other ethnic groups in the sample have only 1 participant. Plus, the ethnic groups are all over the place; "Asian" is a single category despite referring to an entire continent, whereas "Bangladeshi" and "Lebanese" refer to a single country. Likewise, one participant is simply labelled "(Other)". Overall, this seems an imbalanced distribution regardless of diagnosis.

#DIAGNOSIS: With Autism Spectrum Disorder coded as A and Typically Developing coded as B, we have a fairly balanced sample of 29 with ASD and 32 with TD. The interesting thing is whether those two groups are balanced.

#Alright, we can comment later, we're spending too much time on this:

##SEPARATING ASD AND TD into two dataframes: ASD_first and TD_first

ASD_first <- first_visit %>% filter(diagnosis == "A")
TD_first <- first_visit %>% filter(diagnosis == "B")

summary(ASD_first)
summary(TD_first)

#The two groups are fairly balanced in terms of ethnicity and gender (although by far most of the TD kids were white). ASD kids have lower mean socialization scores than TD kids, though similar scores in nonverbal IQ and only slightly lower verbal IQ scores. Initially, the mean MLU of both mother and children are similar across groups, and so are mean types of mother and child across groups. Interestingly, however, the age-matching isn't quite right. ASD kids have a mean age of 33 months compared to 20 for TD kids, and a much larger spread of age (19-42 compared with 18-24). In short, on average the ASD kids are older.


#Alternative (but slower) way to characterize amount of diagnoses
nrow(subset(first_visit, diagnosis == "A"))
#29 have ASD
nrow(subset(first_visit, diagnosis == "B"))
#32 are TD

#It could also be interesting to see if ages were balanced gender-wise. We can do this in the following way:
gender_mean_age <- first_visit %>% group_by(gender) %>% summarize(n(), mean_age = mean(age, NA.RM = T))
View(gender_mean_age)
#Interestingly, as far as gender is concerned, the mean age is similar. This may be reflected in the fact that the ASD kids are systematically older, and there is a similar amount of girls and boys in each diagnosis group; thus age averages for gender is similar.

#We can do t-tests of nonverbal IQ and verbal IQ at visit 1
nvIQ_test <- lm(nonverbal_IQ ~ diagnosis, data = first_visit)
summary(IQ_t_test1)
#No significant difference of nonverbal IQ at visit 1 between diagnosis group.
#Now for verbal IQ:
vIQ_test <- lm(verbal_IQ ~ diagnosis, data = first_visit)
summary(IQ_t_test2)
#Non-significant difference, but certainly larger. And again, it is worth noting that the ASD kids are generally older, so relative to their age, they should be different. One could look for an interaction effect here.

#Perhaps we should try including age as a predictor, and also include random slopes and intercepts? All of this is just exploring the data, since this isn't my experiment so by definition I cannot do planned comparisons:

nvIQ_test2 <- lm(nonverbal_IQ ~ diagnosis*age, data = first_visit)
summary(nvIQ_test2)
#Something weird happens to the beta-value for diagnosisB here (I assume it has something to do with setting age to 0. Interestingly, this doesn't appear to have the same effect on ASD kids; perhaps because the relationship in TD intelligence as a function of age is higher than for ASD kids? That seems a likely conclusion.)
#Moreover, it is a testable one. Can I predict the nonverbal IQ of TD kids as a function of their age?

TD_nvIQ <- lm(nonverbal_IQ ~ age, data = TD_first)
summary(TD_nvIQ) #There is a significant slope here. Meaning if you assume that for every month you get an IQ point and you then try to imagine then at 0 months... Yikes. That explains that statistical weirdness at least.

ASD_nvIQ <- lm(nonverbal_IQ ~ age, data = ASD_first)
summary(ASD_nvIQ) #And here the slope is much smaller, explaining why the earlier model assumes TD kids at age 0 are stupid but ASD kids at age 0 are ludicrously much smarter. So much for linear models, guys! Still, this is interesting. I showed a relationship between age and nonverbal IQ; the reason they're balanced (one might propose) is that the ASD kids are older. If you allow the TD kids to grow older you start to see them become smarter.

#There may well be numerous statistical problems with my model, but hey, it seems to show what I predicted at least. Huzzah.


```

The sample as a whole included mostly young (<20) white males. The two groups compared to eachother, however, are fairly balanced in terms of gender (in terms of ethnicity, nearly all TD kids were white).
ASD kids have lower mean socialization scores than TD kids, though similar scores in nonverbal IQ and only slightly lower verbal IQ scores. Both differences (without accounting for age difference) are non-significant. Initially, the mean MLU of both mother and children are similar across groups, and so are mean types of mother and child across groups. Interestingly, however, the age-matching isn't quite right. ASD kids have a mean age of 33 months compared to 20 for TD kids, and a much larger spread of age (19-42 compared with 18-24). In short, on average the ASD kids are older.

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}
#Plotting children's MLU based on visit, with two different slopes based on diagnosis,
CHI_MLU_plot <- ggplot(all_data, aes(visit, CHI_MLU, fill = diagnosis, colour = diagnosis))
CHI_MLU_plot + geom_smooth(method = "lm") + geom_point(position = "jitter")
#Perhaps something similar could be done for every single kid:
snowflake_MLU_plot <- ggplot(all_data, aes(visit, CHI_MLU, colour = child_ID))
snowflake_MLU_plot + geom_smooth(method = "lm", fill = "transparent") + geom_point(position = "jitter")
#Colourful but not particularly useful. Oh well, it's a little bit informative.


#Finally we get around to making a statistical test in order to test Hypothesis 1. We will make two models, one with and without an interaction term of visit and diagnosis. Then we will compare the two:
h1_lm <- lmer(CHI_MLU ~ visit + diagnosis + (1|child_ID) + (0 + visit|child_ID), data = all_data)
h1_interact <- lmerTest::lmer(CHI_MLU ~ visit*diagnosis + (1|child_ID) + (0 + visit|child_ID), data = all_data)
#Convergence issue. Ouch.

summary(h1_interact)
```

How would you evaluate whether the model is a good model?

```{r ex2 evaluate, include = FALSE}
#Comparing with the null model
anova(h1_lm)

#Comparing with a different model
anova(h1_interact, h1_lm)

#Comparing THAT with the null model
anova(h1_interact)

#We certainly have significant results of visit*diagnosis, but how do I assess whether it is good? 

```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better.

```{r ex2 growth curve, include = FALSE}

```

Exciting right? Let's check whether the model is doing an alright job at fitting the data. Plot the actual CHI_MLU data against the predictions of the model fitted(model). 

```{r}
#I assume by this he means plot both with points and regression lines?
```

Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your model's predictions (and some comments on whether the predictions are sensible)

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}

#Plotting/inspecting the data
parental_plotting <- ggplot(ASD_data, aes(visit, MOT_MLU, fill = diagnosis, colour = diagnosis))
parental_plotting + geom_point(position = "jitter") + geom_smooth(method = "lm")


h2_lm <- lmerTest::lmer(MOT_MLU ~ visit + diagnosis + (1|child_ID) + (0 + visit|child_ID), data = ASD_data)
h2_interact <- lmerTest::lmer(MOT_MLU ~ visit*diagnosis + (1|child_ID) + (0 + visit|child_ID), data = ASD_data)

#
summary(h2_interact)

```

Parent MLU is affected by ... but probably not ...
[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Kenneth


```{r ex4, include = FALSE}
#Time for: Ridiculously complicated models, and pseudo-systematic data exploration! What could go wrong?

#The general idea will be to add onto the existing model from before, and seeing if additional variables improves the model's predictive power. For instance, does verbal IQ interact with the previous effect?

h1_plus_vIQ <- lmerTest::lmer(CHI_MLU ~ visit*diagnosis*verbal_IQ + (1|child_ID) + (0 + visit|child_ID), data = all_data)
summary(h1_plus_vIQ)

anova(h1_interact, h1_plus_vIQ)

#Verbal IQ is very predictive of MLU when all other variables (diagnosis and the like) are zero, which makes sense. There are no significant interaction terms; seemingly verbal IQ doesn't mean the child develops faster or slower with visits...
#Ya know, it's kind of a ridiculous variable to include since verbal IQ and MLU might as well be the same thing.

#Trying socialization as a variable!
h1_plus_soc <-lmerTest::lmer(CHI_MLU ~ visit*diagnosis*socialization + (1|child_ID) + (0 + visit|child_ID), data = all_data)
summary(h1_plus_soc)
#Meh.

```

In addition to ..., the MLU of the children is also correlated with ...
Using AIC / nested F-tests as a criterium, we compared models of increasing complexity and found that ...

[REPORT THE RESULTS]